Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-23 write-ups/FSE/paper.tex
Add overview figure
did a little bit of work

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.
\ADComment{Tyler: briefly state that this still occurs on Pascal, despite it having support for preemption.}

@@

@@

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that GPU vendors (1) will
not commit to the occupancy-bound execution model they currently
implement, for the above reasons, yet (2) will not guarantee fair
scheduling due to the runtime cost of preempting workgroups in a fair
manner.  Vendors instead wish to retain the essence of the simple
occupancy-bound model, supporting preemption only in key special
cases.

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
%\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\vspace{-8mm}
\end{table}

@@

\subsection{Comparison with Preemption on NVIDIA GPUs}\label{sec:nvidiacomparison}

>>>>

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.
\ADComment{Tyler: briefly state that this still occurs on Pascal, despite it having support for preemption.}
\TSComment{I added this but actually think it fits better a little bit later (marked with a comment). See what you think}

@@

\begin{figure}[t]
\centering
\includegraphics[width=\columnwidth]{overview.pdf}
\caption{Cooperative kernels can flexibly be resized in order to let
other tasks (e.g. graphics) access GPU ressources}
\label{fig:overview}
\end{figure}

@@

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that GPU vendors (1) will
not commit to the occupancy-bound execution model they currently
implement, for the above reasons, yet (2) will not guarantee fair
scheduling due to the runtime cost of preempting workgroups in a fair
manner.  Vendors instead wish to retain the essence of the simple
occupancy-bound model, supporting preemption only in key special
cases.
%
\TSComment{added note about Pascal experiment here. I wonder if there's
a better way to glue these paragraphs together? That is, above it provides
two issues vendors have with forward progress and below, we show a concrete
example of both issues in Nvidia's latest chips}
%
For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{??}. Despite this, Nvidia's
documentation still does not provide concrete forward progress
properties (i.e. vendors will not yet commit to providing these
properties). Furthermore, we were able to run some small experiments
on a Pascal GPU (GTX Titan X) and observed that the GPU still adhered
to the occupancy-bound execution model. That is, a kernel containing a
global barrier executed successfully when run with 56 workgroups, but
deadlocked when run with 57 workgroups. This is an example of a vendor
supporting preemption, yet not providing full fairness properties.

@@

\begin{table}[t]
\normalsize
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
%\rowcolor{Gray1}
%bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
%\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\vspace{-8mm}
\end{table}

@@

\subsection{Preemption on NVIDIA Comparison}\label{sec:nvidiacomparison}
